Towards a free high quality powder X-ray diffraction dataset

### 2024-04-10 paper/01_motivation.tex
Highlighted reference library problem

% On Rietveld refinement
The objective of powder x-ray diffraction experiments is to extract as much information about the crystal structure
of the sample as possible, or if the sample is multi-phase, also identify the phases present in the sample and quantify
their fractions.
This is done through a technique called Rietveld refinement: A working model of the crystal structure, starting
from an initial guess provided by the operator of the analysis software, is assumed and the corresponding diffractogram
is simulated.
The parameters of the working model of the sample are then adjusted to minimize the difference between the simulated
diffraction pattern and the one observed in experiment.
This workflow often works well, since usually the operator of the experiment will be able
to provide an at least somewhat accurate initial guess of the crystal structure.

However, employing manually operated Rietveld Refinement is not without issues and it is certainly not
scalable to the degree that would be required to keep up with advances
in throughput and efficiency in other steps of the experimentation pipeline, if just due to the manual effort involved.

@@

While the ICSD contains a large number of structures spanning many different
classes of materials, it still falls short in size, distribution, and generality
compared to the datasets used to train very large state-of-the-art models of
other fields such as computer vision.
%for image classification (ImageNet
%challenge\supercite{russakovskyImageNetLargeScale2015}). 
Furthermore, the ICSD
database is highly imbalanced with respect to space groups, as can be seen
in the histogram in Figure~\ref{fig:histogram_dist_spgs}. This makes the classification of
space groups more difficult, as shown and discussed by
\citeauthor{zalogaCrystalSymmetryClassification2020}
\supercite{zalogaCrystalSymmetryClassification2020}. The ICSD also contains a
limited number of different structure types that may not adequately represent
the crystal structures analyzed in future experiments.

To overcome these shortcomings, we propose to train machine learning models on
diffractograms simulated from synthetic crystal structures randomly generated
based on the symmetry operations of the space groups. This makes it possible to
train on structures with new structure types not present in the ICSD. We used
the crystals from the ICSD only to determine vague statistics guiding the random
generation and for calculating the test accuracy. Our approach goes one step
further than classical data augmentation by fully detaching itself from the
individual entries in the ICSD database.
\replaced{The generated synthetic crystals form a training dataset that includes stable ICSD crystal structures, unstable crystal structures,
but also stable structures that are not yet present in the ICSD. By training a model on the full dataset, we can also expect improvements on the unknown stable crystal structures.}{It is our hypothesis that when trying
to solve more general tasks concerning the extraction of structural details from
powder diffractograms, the crystals from which the training dataset is simulated
do not need to be stable or chemically viable.} \replaced{Furthermore, we}{We} propose viewing the problem
\deleted{simply} as a mathematical task of getting back some of the real-space information
leading to given powder X-ray diffractograms. \added{Therefore, even the unstable structures included in our generated dataset will help to learn to classify the stable structures.}

Here, we applied this approach to the classification of the crystal symmetry,
namely the space group. The space group is usually one of the first structural
pieces of information needed after synthesizing a new material. This task is
well-suited to showcase the strengths of using a synthetic dataset and to
benchmark it. We further show the results of using our methodology to infer space group labels of an experimental dataset.

We embedded our synthetic generation algorithm in a framework with
distributed computing capabilities to generate and simulate diffractograms on
multiple nodes in parallel using the \emph{Python} library \emph{Ray}
\supercite{moritzRayDistributedFramework2018a}. In contrast to the traditional
approach of generating a simulated dataset before training, we used this
distributed computing architecture to build an infinite stream of \replaced{synthetically}{randomly}
generated and simulated diffractograms to perform batch-wise online learning.
This increases the generalization performance, eliminates the problem of
overfitting, and allows very large models to be trained.

>>>>

% On Rietveld refinement
The objective of powder x-ray diffraction experiments is to extract as much information about the crystal structure
of the sample as possible, or if the sample is multi-phase, also identify the phases present in the sample and quantify
their fractions.
This is done through a technique called Rietveld refinement: A working model of the crystal structure, starting
from an initial guess provided by the operator of the analysis software, is assumed and the corresponding diffractogram
is simulated.
The parameters of the working model of the sample are then adjusted to minimize the difference between the simulated
diffraction pattern and the one observed in experiment.
Critically, this process does not generally guarantee to guide the working model towards a global minimum of the difference
in diffractograms and hence arrive at an accurate model of the sample.
In practice the refinement process may only guide the working model towards the nearest local minimum
starting from the initial guess.
Hence, the result of the refinement procedure is in general only as good as the initial guess it was provided with.
Still this workflow often works well, since usually the operator of the experiment will be able
to provide an at least somewhat accurate initial guess of the crystal structure.

However, employing manually operated Rietveld Refinement is not without issues.
It is certainly not scalable to the degree that would be required to keep up with advances
in throughput and efficiency in other steps of the experimentation pipeline, even just due to the manual effort involved.
But beyond the lacking scalability of this workflow we also wish to highlight the following issue:
Providing the analysis software with an initial guess of the structure requires consulting a reference library
of crystal structures with corresponding simulated diffractograms and
identifying a crystal structure that is likely to be similar to that of the investigated sample.
Since the number of chemically viable crystal structures is infinite, any such library will always be incomplete.
The best that can be done to remedy this is to continually update such a reference library both on the side
of the distributor and the user, which in itself represents an undesirable overhead.
Especially for novel material discovery it may be impossible to find a sufficiently accurate match in the reference
library to apply Rietveld Refinement successfully.

Machine learning is a very promising candidate to eliminate both the need for manual inputs and the need for reference
libraries in powder x-ray diffraction analysis.

@@
